Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-20 submission/manuscript.Rmd
table notes
drafting discussion

- testis genes are germline genes, germline gene list
	- While a select few tissue-enriched DEGs had known roles in the brain, the majority were testis genes with no known function outside the germline. 
	- Studies on chromatin regulators in germline gene repression has focused on specific genes 
	- Curated a list of germline-enriched genes to robustly assess the types of genes dysregulated and KDM5C's role

- KDM5C represses germline genes during early embryogenesis
	- Found this dysregulation in cellular identity begins during _Kdm5c_-KO embryogenesis
	- EpiLC express early regualtors, Brain late sperm - program seems to be continuing across development
	- Importantly, even though germline genes are misexpressed, markers of EpiLC differentiation and EpiLC morphology is unimpaired. 
		- Epiblast can become primoridal germ cells
		- Otx2 is expressed in EpiLCs and is known to repress PGC identity. It's properly expressed in _Kdm5c_-KO EpiLCs, further supporting they aren't just becoming PGCs https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/ 
		- Germline Program happening in the background of typical development.

@@

- We globally characterized KDM5C's repression of germline genes during early embryogenesis and in mature neurons. 
	- In line with previous work, we found KDM5C did not regulate germline genes in neurons. 
	- KDM5C-bound genes in EpiLCs were enriched for germline ontologies, suggesting a major role of KDM5c during embryogenesis
	- While KDM5C directly binds some germline genes, including _Dazl_, many of the genes dysregulated in _Kdm5c_-KO were not direct targets. 
		- Stra8 notable exception
		- Brain genes, including late-stage spermatogenesis genes
		- Suggests germline genes can be dysregulated direct and indirect of KDM5C resgulation
		- Further supports germline programs can be ectopically activated

- Germline genes typcially decomissioned during ESCs and EpiLC differentiation. 
	- Initially repressive histone marks are placed in ESCs, then DNA methylation is placed in EpiLCs.
	- Dysregulation of germline genes later in life and in EpiLCs indicates KDM5C is involved in this processes, but unclear when. 
	- Recently in ESCs KDM5C's catalytic acivity was found to be unnecessary for dazl suppression.  
	- Some evidence that KDM5C is important for DNAme, as DNAme is reduced at two germ gene promoters in the adult 5CKO hippocampus.
	- We found KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing.

- Altogether, this suggests KDM5C plays a pivotal role in the development of tissue identity, including the soma-germline distinction.

@@

\newpage
# Figures and Tables

>>>>

- testis genes are germline genes, germline gene list
	- While a select few tissue-enriched DEGs had known roles in the brain, the majority were testis genes with no known function outside the germline. 
	- distinguishing between germline and soma a key feature of multicellularity, chromatin regulators are important for this process
	- Studies on chromatin regulators in germline gene repression has focused on specific genes 
	- Curated a list of germline-enriched genes to robustly assess the types of genes dysregulated and KDM5C's role

- KDM5C represses germline genes during early embryogenesis
	- Found this dysregulation in tissue identity begins during _Kdm5c_-KO early embryogenesis
	- EpiLC express early regualtors, Brain late sperm - program seems to be continuing across development
	- Importantly, even though germline genes are misexpressed, markers of EpiLC differentiation and EpiLC morphology is unimpaired. 
		- Epiblast can become primoridal germ cells
		- Otx2 is expressed in EpiLCs and is known to repress PGC identity. It's properly expressed in _Kdm5c_-KO EpiLCs, further supporting they aren't just becoming PGCs https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/ 
		- Germline Program happening in the background of typical development.

@@

- We globally characterized KDM5C's repression of germline genes during early embryogenesis and in mature neurons. 
	- In line with previous work, we found KDM5C did not regulate germline genes in neurons. 
	- KDM5C-bound genes in EpiLCs were enriched for germline ontologies, suggesting a major role of KDM5c during embryogenesis
	- While KDM5C directly binds some germline genes, including _Dazl_, many of the genes dysregulated in _Kdm5c_-KO were not direct targets. 
		- Stra8 notable exception
		- Brain genes, including late-stage spermatogenesis genes
		- Suggests germline genes can be dysregulated direct and indirect of KDM5C resgulation
		- Further supports germline programs can be ectopically activated
		- Talk about different motifs if we do see differences there and if it explains direct vs indirect dysregulation
			- Could also mention PRC1.6/KDM5C binds pcgf6 in ESCs

- We demonstrated KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing of germline genes during the transition from naive to primed pluripotency.
	- Loss of DNAme can last throughout life at at least two gemrline gene promoters (hippocampus barco)
	- KDM5C could be important for DNAme is that KDM5C erases H3K4me3 which can impede CpGme.
		- In support of this, KDM5C is highly enriched at the CpG islands near germline TSS that are methylated in EpiLCs
	- However, Recently KDM5C's catalytic acivity was found to be unnecessary for dazl suppression in ESCs.  
	- Since DNAme is not placed until EpiLC stage, KDM5C's catalytic activity may be required for long-term silencing of germline genes.

- In summary, we demonstrate loss of the X-linked intellectual disability gene KDM5C leads to widespread dysregulation of brain tissue identity, including misexpression of germline genes in the somatic brain. 
	- In EpiLCs, KDM5C directly represses key drivers of germline identity like Dazl, likely through PRC1.6 recruitment and promoting CpG methylation. However, _Kdm5c_-KO also ectopically expresses germline genes activated indirectly, including _Stra8_.  
	- The germline developmental program to some extent continues ectopically during _Kdm5c_-KO development, resulting in aberrant transcription of late stage spermatogenesis genes later in life. 
	- These results define KDM5C's role in the demarcation between soma and gemrline identity and offer a window into potential targets to assess the deletirous ecfects these ectopic genes on neurodevelopment.

## Discussion notes
- Altogether, this suggests KDM5C plays a pivotal role in the development of tissue identity, including the soma-germline distinction.

@@

\newpage
# Figures and Tables
- Supplementary table 1: list of all germline genes. 
	- Columns to include: 
		- KDM5C bound vs not
		- DEG in EpiLC, brain, both, neither (separate columns?)
